---
title: "Citi Bike Ride Duration Analysis"
author: "Manh Tuong Nguyen"
output:
  word_document: default
date: "2024-11-18"
editor_options:
  markdown:
    wrap: 72
    
always_allow_html: yes
---

# Data importation

I first read in the data file and drop all the rows that has N/A values
since the data is very large. Also, for October 2024, the total data has
6 parts so I only use a half of them for optimizing simplicity and
performance.

```{r, include=FALSE}
library(tidyverse)
library(ggplot2)
library(sf)
library(mapview)
```

```{r, include=FALSE}
temp <- list.files(path='/Users/eddie/MATH167R/Data', pattern="\\.csv$", all.files=TRUE, full.names = TRUE) # Create a list of .csv files in the clarified path
myfiles <- lapply(temp, read.csv) # Read all the files using csv and make a list to store all data.frames
data <- do.call(rbind, myfiles) # Bind all the data.frames into one
rm(myfiles) # Erase the list from memory to free space
data <- drop_na(data = data) #Drop all rows with N/A values
data <- subset(data, select =  -c(end_station_id, start_station_id, ride_id)) # Drop columns that contains IDs
data <- data.frame(data) # Ensure the data is stored as data.frame
data$started_at <- as.POSIXct(data$started_at, format = "%Y-%m-%d %H:%M:%S") # Reformat the data as POSIXct date
data$ended_at <- as.POSIXct(data$ended_at, format = "%Y-%m-%d %H:%M:%S")
data <- data[(data$started_at > as.POSIXct("2024-10-01 00:00:00")) & (data$started_at < as.POSIXct("2024-10-14 23:59:59")),] # Ensure the data has rides started from 0:00:00 Oct 10, 2024 to 23:59:59 Oct 14, 2024
```

# Exploratory Data Analysis

## Shape and columns

```{r}
glimpse(data)
```

The cleaned data has 2340229 rows with 10 columns. The data type of all
the columns are well-formatted.

```{r}
data$datetime_minute <- format(data$started_at, "%Y-%m-%d %H:%M") # Extract date-time up to minutes
group_by(data, datetime_minute) |> 
  summarise(count = n()) |> 
  ggplot(aes(x=as.POSIXct(datetime_minute, format='%Y-%m-%d %H:%M'), y=count)) + 
  geom_line(color='blue') + labs(title='Count by hour and minute', x='Timespan', y='Count') + 
  theme_classic()
```

Here I will create the duration data by calculating the differences
between started_at and ended_at

```{r}
data$duration <- round(as.numeric(difftime(data$ended_at, data$started_at, units='secs')))
data$duration <- data$duration / 60
```

```{r}
ggplot(data[data$duration < 60,], aes(x=duration, color=member_casual)) + geom_histogram(bins=60) + labs(title='Historgram of ride duration under 60 minutes', x='Duration (minutes)', y='Count') + theme_classic() 
```

Next, I will create a column for storing the day of the week

```{r}

weekdays <- factor(c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"),
                   levels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

data$weekday <- weekdays[wday(as.POSIXct(data$started_at))]

```

With bins = 60, each bin represent one minute additional to the ride. We
can see that people in NYC most likely rent a bike to use for 4-7
minutes. Let's conduct t-test to see if memebership type has any effect
on renting duration

```{r}
summary(data)
```

### Data Cleaning

The statistical summary shows that the data has outliers. We need to
take care them first before proceeding to the analysis

```{r}
boxplot(data$end_lng)
boxplot(data$end_lat)
```

```{r}
count(data |> filter(end_lat == 0))
count(data |> filter(end_lng == 0))
```

The end lat and end_lon contains coordinations that are 0. Since there
are only 15 entries looking like that, we are going to remove all of
them.

### Outliers in duration

As showing in the histogram of duration above, there are a lot of
outliers in the dataset for duration. My method is to label the
observations with duration is longer than 1.5 \* InterQuartile Range
from Q3.

```{r}
#Calculate the outlier initial value

duration_quantiles <- quantile(data$duration, probs=c(.25, .5, .75), type=1)

iqr <- duration_quantiles[3] - duration_quantiles[1]

duration_outlier_val <- duration_quantiles[3] + 0.25 * iqr

# Filter values

prop_true <- sum(data$duration >= duration_outlier_val) / nrow(data)

cat("Proportion of TRUE:", prop_true, "\n")
```

Since there are only 0.055 of the data contains outliers rides, we
discard those outliers to clean up the dataset.

```{r}
data <- data |> filter(data$duration < duration_outlier_val)
```

```{r}
hist(data$duration)
```

```{r}
boxplot(data$duration)
```

Here I add extract and add a column of the start time (in hour)

```{r}
convert_hour <- function(time) {
  hour <- as.numeric(format(time, "%H"))
  minute <- as.numeric(format(time, "%M"))
  
  return(hour + minute/60) 
}

data$start_time <- sapply(data$started_at, convert_hour)
```

# Subquestion 1: Do trips taken by subscribers differ in duration compared to trips taken by casual users?

```{r}
member_duration <- data[data$member_casual == 'member',]$duration
casual_duration <- data[data$member_casual == 'casual',]$duration
```

For this question, I want to perform t-test to test to confirm if the
trip duration of rides taken by members differ from rides taken by
casual drivers.

Null Hypothesis: the mean of ride duration for member drivers and casual
drivers are not different

Alternative Hypothesis: there is difference in the mean of ride duration
for member and casual drivers

## Assumption check

### Normality

```{r}
for (membership in c('member', 'casual')){
  
  qqnorm(data[data$member_casual == membership,]$duration)
  qqline(data[data$member_casual == membership,]$duration, col='blue')
}
```

The qq plots suggest the data are heavy-tailed. However, we can see that
the middle range of the sample quantile is close to the normal line.
Plus, since the data size is very big, we can use Central Limit Theorem
to conclude the distribution of the mean is normal.

```{r}
mean_and_variance <- function(data) {
  nu <- mean(data, na.rm = TRUE)
  variance <- var(data, na.rm = TRUE)
  cat('Mean: ', nu, ' Variance: ', variance)
}

cat("Member's duration: ")
mean_and_variance(member_duration)
cat("\nCasual's duration: ")
mean_and_variance(casual_duration)
```

So, we can conclude that member's duration and casual's duration satisfy
the normality of mean assumption with the above mean and variance. Also,
they potentially have different means, too.

### Independence

In this dataset, the data was collected individually from mostly all
users in New York City, which may contain the two or more entries come
from a person. However, the trip durations are not depended on each
others since the data was randomly collected and the chance of one bike
ride's duration influence another is very low as it go against the
common sense. So, the observations are independent from each others.

### t-test

Now we have checked all the assumptions, let's proceed with 2 samples
t-test.

```{r}
t.test(duration ~ member_casual, data=data)
```

First of all, t-test's p-value (\< 0.05) indicates that we have
sufficient evidence to reject the null hypothesis. Hence, the average
duration of bike rides of member and casual groups are different. With
that being said, membership has influence on ride's duration.

```{r}
ggplot(data=data[data$duration < 60,], aes(x=duration, color=member_casual, group=member_casual)) + geom_histogram(bins = 60) + theme_classic() + labs(title='Histogram of ride duration under 60 minutes', x='Duration (mins)', y='Count', color='Membership')
```

```{r}
ggplot(data=data[data$duration < 60,], aes(x=duration, fill=member_casual, group=member_casual)) + 
  geom_boxplot(outlier.shape = NA) + labs(title='Box plot of ride duration between casual and member users', x='Duration (in minutes)', fill='Membership') + theme_classic()
```

In particular, casual users tend to use the bike for longer duration
than member drivers. This trend makes sense as people who subscribes to
the membership plan usually commute regularly using bike (bike service
from this program, specifically). If they commute regularly, it is more
likely that the range of the trip is short so that they can save time to
find parking or money on gas. For long trips, it is more likely that the
bikes are used in emergency cases rather than regular commute. Also,
maybe they just want to try out the service. As a result, this insight
explain why membership affect the bike duration. Last but not least,
membership should be included in our linear regression model for
duration.

# Subquestion 2: What is the effect of bike type (electric vs. classic) on the trip duration, controlling for the day of the week?

E-bikes offer an effortless ride with the help of motors, while classic
bikes require riders to use their pure strengths to move the bike. In
New York City, the streets are often crowded, and bicycles have become a
popular means of transportation for many residents, especially for
commuting to work (NYC Department of Transportation, 2022). Due to the
convenience, e-bikes’ rates are significantly higher than that of
classic bikes. So, it is interesting to see how the ride duration is for
each type of ride with the consideration of the day of the week. This
finding will tell about the economic effectiveness of electric bikes
compared to classic bikes.

```{r}
data |> group_by(rideable_type, weekday) |> summarise(count = n())
```

In all the groups, the number of entries are all very large. Hence, we
can use CLT to confirm that the mean distribution of duration among all
groups are normal. Between classic bike and e-bike, the ride duration of
one cannot influence the other as there are no relationship between
them. However, as we seen in the plot of usage timeseries, we can see
that there are seasonality with weekly frequency in the data. So, if we
group the data into day of the week (Monday, Tuesday, etc.), there is a
potential that the data are not dependent. Thus, the setting must be
comparing e-bike and classic bike groups 7 times corresponding to 7 days
a week.

Null hypothesis: There are no different in trip's average duration
between ride type - weekday groups Alternative hypothesis: There exist a
ride type - weekday group that has difference trip's average duration
compare to others

```{r}
results <- list()

# Loop through each weekday
for (day in weekdays) {
  # Filter the data for the current day
  tmp <- data[data$weekday == day, ]
  
  # Run t-test
  test <- t.test(duration ~ rideable_type, data = tmp)
  
  # Save the summary of the t-test in the results list
  results[[day]] <- test
}

print(results)
```

Among all the test results, the p-value are always less than 0.05
suggesting that we can reject the null hypothesis and conclude that the
trip's average duration between classic bike and electric bike are
diffrent controlling by day of the week. Furthermore, the mean of ride
duration of e-bikes is always higher than classic bikes. Hence, it means
that the e-bike is mainly picked if people want to go for a long ride
while classic bikes are more suitable to short rides. Let's plot them
all on a plot to identify the weekly trend.

```{r}
ggplot(data = data, aes(x=weekday, y=duration, colour = rideable_type)) + geom_boxplot(outlier.shape = NA) + labs(title='Box plot of ride duration between casual and member users', x='Duration (in minutes)', fill='Ride type') + theme_classic()
```

Averagely, the duration of rides of both classic bikes and e-bikes peak
during weekend days. This can be explained because people who want to
use the bikes as a way to enjoy the fresh air or exercise after a busy
week contribute to the surge of demand. As a result, Citi Bike should
have a routine distribution of bike, meaning that they should store
bikes in warehouse during the week and release them fully during the
weekend to minimize the chance of bikes become damaged due to frequent
usages. Also, we can conclude that rideable type and day of the week are
significant predictor for the duration linear regression model.

# Subquestion 3: How does ride duration differ at different starting locations?

For this question, I want to check which region tend to have longer
rides compare to others at different time a day, in other words, the
ranking of demand at different regions. Knowing this will help decision
makers gain more insights on where to ship newly charged bikes or
schedule more frequent maintenance to meet the demand.

First, I created the regions using grouping. I will focus on 5 different
regions: Downtown Manhattan, Central Park, Uptown New York + The Bronx,
Brooklyn, and Queens. Those are the areas that represented in the
dataset. The New York main island part was separated into 3 parts to
increase classification accuracy as it is very wide and dense.

Due to the size of the data set, it will take a lot of time to use API
to map the start stations coordination to the exact region. So, I
created 5 central coordinates corresponding to 5 big areas and performed
clustering using Euclidean distance. In the plot, we can see that there
are some mismatch in Brooklyn. It is because of the points that we used
as the center. However, the grouping shows the distinction between the
areas so we are good to go.

```{r}
library(tidygeocoder)


locations <- data[, c('start_lat', 'start_lng')]

find_nearest_region <- function(lat, lon, regions) {
  # Calculate great-circle distances using the Haversine formula
  distances <- sqrt((regions$lat - lat)^2 + (regions$lng - lon)^2)
  
  # Find the region with the minimum distance
  nearest_region <- regions[which.min(distances), ]
  
  # Return the nearest region's name
  return(nearest_region$region)
}

# Example usage
regions <- data.frame(
  region = c(
    "Downtown Manhattan", 
    "Brooklyn Downtown", 
    "NY Uptown + The Bronx", 
    "Central Park",
    "Queens (Long Island City)"
  ),
  lat = c(
    40.732724,  # Downtown Manhattan
    40.689089,  # Brooklyn Downtown
    40.820514,  # The Bronx
    40.778834, # Central Park
    40.762762  # Queens (Long Island City)
  ),
  lng = c(
    -73.991779, # Downtown Manhattan
    -73.957245, # Brooklyn Downtown
    -73.935164, # The Bronx
    -73.973304, # Central Park
    -73.920252 # Queens (Long Island City)
  )
)

locations$regions <- mapply(find_nearest_region, locations$start_lat, locations$start_lng, MoreArgs = list(regions))
```

```{r}
data$start_region <- locations$regions

plot_dat <- locations |> 
  distinct(start_lat, start_lng, regions) |> 
  st_as_sf(coords = c("start_lng", "start_lat"), crs = 4326)

mapview(plot_dat, cex = 3, label = "regions", zcol = "regions", col.regions = c("blue", "green", "yellow", "red", "orange"))

```

### Assumption check

The plan is to use ANOVA 5 times to compare the average ride duration at
5 regions. Since, the ride duration of one region will not influence the
usage at others, we can safely assume the independence assumption is
satisfied. Since the data size is large, we can use Central Limit
Theorem to conclude that the distribution of average duration in all the
5 regions are normal.Now, we will create the usage data and check
normality

```{r}
cat('Mean and variance of different regions:\n')
for (region in regions$region) {
  plot_data <-  data %>% filter(start_region == region) %>% drop_na() 
  plot_data <- plot_data$duration
  cat(region, ': ')
  mean_and_variance(plot_data)
  cat('\n')
}

```

Next up, we can see that there is potential different in the means of
all 5 groups. Plus. the variance of all groups are very close to each
other. Hence, the data passed the potential mean different and
homogeneity assumption

So, the data already passed all the assumptions, we will proceed with
the ANOVA analysis:

### ANOVA

Alpha = 0.05

Null hypothesis (H0): The mean of duration at all 5 regions are similar
to each other Alternative hypothesis (HA): There is at least one region
that has average ride duration different from other regions.

```{r}
model <- aov(duration ~ start_region, data=data)
summary(model)
```

Since ANOVA result in p-value \< 0.05, we have sufficient evidence to
reject the null hypothesis and conclude that there is at least one
region that has average ride duration different from other regions. In
addition, I will perform Tukey's post hoc test to identify the
differences between groups in pair.

```{r, echo=TRUE, fig.width=9}
TukeyHSD(model, conf.level = 0.95)
```

The result of Tukey told us that all the groups are distinct using
p-value. For the Queens - NY Uptown + The Bronx pair, since p-value is
not under 0.05, we conclude that the 2 groups are not different in term
of ride duration. This result means the average duration of rides taken
in 5 major areas of New York City are different. Let's corporate
boxplots to compare the groups to give the ranking of average duration

```{r}
# Create the plot
ggplot(data, aes(x = start_region, y = duration)) +
  geom_boxplot() +
  labs(title = "Boxplot of Duration by Start Region", 
       x = "Start Region", y = "Duration") +
  theme_classic()

# Adjust the plot size when saving
#ggsave=("boxplot.png", plot = p, width = 20, height = 6, units = "in")
```

Based on all the results, we can see that the average ride duration in
Central Park is longest. Runner up is Brooklyn Downtown then Downtown
Manhattan. And the final place are both NY Uptown + The Bronx and
Queens. Using this insight, we can conclude two things:

-   As the result from sub-question 2, Citi Bike should arrange the most
    e-bikes in Central Park and promote their use, as customers there
    are more likely to go for long trips and find them more appealing,
    while reducing the number of classic bikes. The classic bikes should
    be put at Queens and NY Uptowns + The Bronx areas because of the low
    average duration. For short rides, it is more likely that people use
    it for short commute or exercise. Hence, classic bikes will be more
    practical for users in those 2 regions

-   To improve customer satisfaction, Citi Bike can focus the
    maintenance team into servicing the bikes in areas with high average
    ride duration as the bikes there will more likely to travel at long
    distance so it will more likely to have issues than in regions with
    low average ride duration.

In general, region of start_station is a significant predictor for
duration in our linear regression model.

# Main question: What factors significantly influence the duration of Citi Bike trips in New York City Central Park?

At first, I wanted to do the whole city scaled, but it is tedious to
diagnose the linear model with such large dataset. Hence, I will focus
the most on the Central Park region as this is a unique location where
bikes can be used for various purposes (commuting, exercising, etc.). We
can possibly see the effect of weekday and start time here using the
variability in purpose of bike use.

```{r}
data$start_lat <- scale(data$start_lat, scale = FALSE)
data$start_lng <- scale(data$start_lng, scale = FALSE)

data |> group_by(start_region) |> summarise(count = n())
tmp <- data[(data$start_region == 'Central Park'),]
tmp
```

Firstly, as stated before, since the data was collected using random
sampling as each ride does not relate to each other. So, we can safely
confirm the data is independence.

So, let's begin with assumption checking by first fitting the model

```{r}
model <- lm(log(duration) ~ member_casual + (start_lat : start_lng) + rideable_type * (start_time * weekday) , data=tmp)
plot(model)
```

```{r}
ggplot(data = data.frame(fitted = fitted(model), residuals = residuals(model)),
       aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.3) + 
  geom_hline(yintercept = 0, color = "red") +
  labs(x = "Fitted Values", y = "Residuals", title="Residual Plot")
```

For the residual, I used log transformation to reduce the skewness of
duration. If we look at the QQ-plot, the upper part turn away from the
qq line, which shows that there is heavy skewness in one side of the
errors. In contrast, the lower part align very close to the normal line.
We cannot use CLT here as it only applies to the sample mean not the
individual data points. So, using this result we can say the residuals
are not completely normally distributed. Also, this means that the model
can be over-performed comparing to using the original duration data.
Moreover, the prediction will change to log-scaled duration. If the
model perform well, we need to perform e$^x$ transformation to get the
exact value for duration.

By applying dim on points in the residual plot with low appearances (low
alpha on a point, so points with more appearances will be darker), we
see that there is no general pattern. The scatter points cluster around
-0.5 for major of the fitted values. Thus, we can conclude that the
error has constant variance This indicates that there is some
correlation between at least one predictor and the dependent variable.
However, since the shape of the residual plot is not a horizontal band,
the linearity assumption between duration and the predictors is not
satisfied.

In the residual vs leverage plot, we don't see any observations out side
any red dashed line, hence, there are not any influential points in the
dataset.

In general, the assumptions for the linear regression as almost
satisfied except the linearity and normality of residuals. So, the
result may not be correct.

### Regression Model Result

```{r}
summary(model)
```

```{r}
print(paste('MSE:', round(mean(residuals(model)^2), 3)))

mod_fits <- fitted(model)
my_df <- data.frame(actual = tmp$duration, fitted = mod_fits)
ggplot(my_df, aes(x = fitted, y = actual)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, col = "red") + 
  theme_bw(base_size = 15) +
  labs(x = "Fitted values", y = "Actual values", title = "Actual vs. Fitted") +
  theme(plot.title = element_text(hjust = 0.5))
```

#### Formula:

To create the linear regression model for the duration at stations in
Central Park region, I used membership, start location coordinate and
start time (in hour), day of the week, and ride type. Since we did not
investigate the relationship between membership and other variables, I
let it be the standalone predictor (no interaction). The same applied
for start station coordinate as we have not investigated its
relationship with weekday or start time. Hence I did not include the
interaction for this variable. For ride type, we saw that it varied
among day of the week. So, we will include that interaction and improve
the detail by including ride type interaction with start time in the
formula to represent the purposes of bike usage. In general, our formula
is:

`log(duration) ~ member_casual + (start_lat : start_lng) + rideable_type * start_time * weekday`

Noted that I have performed grid search by trying different combinations
of all the variables and using F-statistics to indicate the good of
fitment, Mean adjusted R-squared (R2) and Mean Squared Error (MSE) to
indicate the accuracy. The above formula is the best one. Noted that
although high F-statistics is good, we have to beware of extremely high
values as it indicates that the model is over-fit if R2 is low and MSE
is high. Moreover, I also performed mean centering to the start
coordinate of start station for better interpretation. This means that
the intercept will account for cases where the station is in the center
of the Central Park region cluster. Then the other station will be
represented by the coordinate different from the center (not distance).

#### Model fitment:

First of all, F-statistics of 290.7 with p-value \< 0.05 shows that this
model is well-fitted. The MSE of 0.391 is relatively small comparing to
the range of values (1-3), however, it is not good enough. Those 2
metrics show that the model is balanced in terms of error and prediction
accuracy, which mean it may not be over-fit or under-fit. Looking at
adjusted-R2, we can see that the value is around 0.024. It means that
this multivariate linear regression model can only explain 2.4% of
log-transformed duration. This is extremely bad already. If we apply
$e^x$ transformation to revert to original scale, the performance get
worse. Hence, this model is not suitable to predict the ride duration of
stations in Central Park region.

#### Conclusion and future questions

By checking all the evaluation metric, we see that the model cannot
represent the characteristic of ride duration in Central Park region.
Moreover, not all the assumptions are met. Hence, the coefficients and
p-value of all the predictors are not correct. Hence, we cannot identify
the most influential factor to ride duration using this approach.
However, it is possible to model the ride duration as the results from
the three sub-questions show that the difference in ride types,
membership, weekday, and start station location can result in different
values of ride duration. Hence, we can use other approaches such as
analyzing and comparing feature importance from Decision Tree based
algorithms to answer this question. Moreover, we can also use other
Machine Learning algorithms create a predictive model for ride duration.
This model will be very helpful for Citi Bike decision makers to assess
and decide "which station should be upgraded to meet the demand of
customer?", or "where should the new station locate?". Also, we can
perform analysis on the year-scaled data to analyze the usage of
stations and corporate that information to the above analysis to
integrate more in-depth information to better answer those questions.

About the dataset, since this is a big data problem, my machine cannot
handle it properly so I had to limit the range of data to only half a
month. Different results might appear if we analyze those questions
again on the data that capture a wider time-span. It will require big
data capable tools like Spark and a stronger machine to handle.

In conclusion, we found out that:

-   People who subscribes to the Citi Bike membership tends to have make
    shorter rides compare to people who do not.

-   E-bikes is more preferred for longer trips and classic bikes is the
    most popular choice for short trips.

-   The ride duration varies at different regions in New York City.

Using the above findings, decision makers can change the operation to
better serve the customers and reduce the chance of damaging the
vehicles by having a better distribution plan.
